---
title: "Plotting gASCA results with ggplot and tidyverse"
author: "Pietro Franceschi"
output: github_document
---

## Introduction

In the ASCA Decomposition of Synthetic Count Data vignette we showed how to run and visualize an ASCA decomposition relying on base graphics.
However, visualizations are often performed using `tidyverse` and `ggplot`. 

The aim of this article is to show how to visualize the results of an ASCA decomposition using these packages. As in the previous case we will use the synthetic data made available in the package:


```{r}
## load the libraries
library(gASCA)
library(tidyverse)
library(patchwork)   ## to combine the ggplots in complex figures

library(ggrepel)


## load the dataset
data("synth_count_data")
```

The following plot shows the distribution of the counts for variables `s1` and `s7` as a function of the design factors:

```{r fig.height=3, fig.width=8}
tibble(synth_count_data$design,
       data.frame(synth_count_data$counts)) %>% 
  pivot_longer(starts_with("s")) %>% 
  filter(name %in% c("s1","s7")) %>% 
  ggplot() + 
  geom_point(aes(x = time, y = value, col = treatment),
             position = position_jitterdodge(dodge.width = 0.4, jitter.width = 0.05), pch = 19, size =2, alpha = 0.5) + 
  scale_color_brewer(palette = "Set1", name = "Treatment") + 
  scale_y_log10() + 
  facet_wrap(~name, scales = "free") +
  xlab("Time") + 
  theme_bw() + 
  theme(aspect.ratio = 0.5)

```

The effect of time on variable `s1` is clear. The presence of an interaction between time and treatment is instead visible in the `s7` profile. 

## Visualize the results of the decomposition and of the validation

### Run the decomposition

```{r}
## perform the decomposition
asca0 <- ASCA_decompose(
  d = synth_count_data$design,
  x = synth_count_data$counts, 
  f = "time + treatment + time:treatment",
  glm_par = list(family = poisson()))

```

### Run the validation

```{r include=FALSE}
## perform the permutation based approach
asca0_validation <- ASCA_permutation(asca0)
```

The following plot shows the results of the decomposition and the validation in terms of L2 norm of the terms, the variable importance and the univariate pseudo-R2

```{r }
figure2_term_val <- asca0_validation$L2_qt %>% 
  as_tibble(rownames = "term") %>% 
  add_column(actual = asca0$terms_L2) %>% 
  mutate(term = factor(term, levels = term)) %>% 
  ggplot() + 
  geom_col(aes(x = term, y = actual), fill = "steelblue") +
  geom_col(aes(x = term, y = value), fill = "darkred") +
  ylab("term-L2") + xlab("") + 
  theme_bw() +
  theme(axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8))
  

figure2_varimp_val <- asca0$varimp %>% 
  as_tibble(rownames = "var") %>% 
  pivot_longer(-var) %>% 
  left_join(asca0_validation$varimp_qt %>% 
              as_tibble(rownames = "var") %>% 
              pivot_longer(-var, values_to = "value_qt")) %>% 
  mutate(var = factor(var, levels = unique(var))) %>% 
  mutate(name = factor(name, levels = unique(name))) %>% 
  ggplot() + 
  geom_col(aes(x = var, y = value), fill = "steelblue") +
  geom_col(aes(x = var, y = value_qt), fill = "darkred") +
  facet_wrap(~name, scales = "free", nrow = 1) +
  ylab("Vecror Norm") + xlab("") + 
  theme_bw() +
  theme(axis.text.x = element_text(size=8, angle = 45, vjust = 1, hjust=1),
        axis.text.y = element_text(size=8))


figure2_R2_val <- asca0_validation$R2_qt %>% 
  as_tibble(rownames = "var") %>%  
  add_column(actual = asca0$pseudoR2) %>% 
  mutate(var = factor(var, levels = unique(var))) %>% 
  ggplot() + 
  geom_col(aes(x = var, y = actual), fill = "steelblue") +
  geom_col(aes(x = var, y = value), fill = "darkred") +
  ylab("pseudo-R2") + xlab("") + 
  theme_bw() +
  theme(axis.text.x = element_text(size=8),
        axis.text.y = element_text(size=8))
  
```


```{r fig.height=5, fig.width=9, message=FALSE, warning=FALSE, dpi=250}

## Define the layout
layout <- "
ABB
CCC
"

## prepare the figure
figure2_term_val + 
  figure2_R2_val + 
  figure2_varimp_val +   
  plot_annotation(tag_levels = 'a') + 
  plot_layout(heights = unit(c(3,4.5),"cm"), widths =  unit(c(6,7),"cm"), design = layout)


```

## Visualization of the decomposition terms

### Perform the decomposition

```{r}
asca0_svd <- ASCA_svd(ASCA_trim_vars(asca0$decomposition, keep = 1:10))
```


### Visualize screeplots, LVs and loadings


#### Screeplot

```{r}
## here we make a scree plot of the three decomposition terms
screeplot <- map(asca0_svd, ~ .x$sdev %>% as_tibble(., rownames = "PC")) %>% 
  enframe(name = "term", value = "data") %>%
  # mutate(data = map(data,~ .x %>% mutate(value = (value/sum(value))*100))) %>% 
  unnest(cols = c(data)) %>% 
  mutate(PC = factor(PC, levels = unique(PC))) %>% 
  mutate(term = factor(term, levels = c("time","treatment","time:treatment"))) %>% 
  filter(PC %in% as.character(1:6)) %>% 
  ggplot() + 
  geom_col(aes(x = PC, y = value^2), fill = "steelblue") + 
  facet_wrap(~term, scales = "free", ncol = 3) + 
  ylab("Variances") + 
  xlab("LV") + 
  theme_bw() + 
  theme(aspect.ratio = 0.4)

```

#### LVs

```{r}
LV_plot <- cbind.data.frame(time_LV1 = asca0_svd$time$x[,1],
                 time_LV2 = asca0_svd$time$x[,2],
                 treatment_LV1 = asca0_svd$treatment$x[,1],
                 "time:treat_LV1" = asca0_svd$`time:treatment`$x[,1], 
                 "time:treat_LV2" = asca0_svd$`time:treatment`$x[,2]) %>% 
  mutate(across(everything(),~round(.x,2))) %>%     
  add_column(synth_count_data$design) %>% 
  pivot_longer(contains("LV")) %>% 
  unique() %>% 
  mutate(name = factor(name)) %>% 
  mutate(name = fct_relevel(name, "treatment_LV1", after = 2)) %>% 
  ggplot() +
  geom_point(aes(x = time, y = value, col = treatment), alpha = 0.5, size = 3) + 
  geom_line(aes(x = time, y = value, col = treatment, group = treatment, lty = treatment)) + 
  facet_wrap(~name, ncol = 1, scales = "free_y") + 
  scale_color_brewer(palette = "Set1") +
  theme_bw() + 
  theme(aspect.ratio = 0.3, legend.position = "bottom")


```

#### Loadings

```{r}
loading_plot <- cbind.data.frame(time_LV1 = asca0_svd$time$rotation[,1],
                 time_LV2 = asca0_svd$time$rotation[,2],
                 treatment_LV1 = asca0_svd$treatment$rotation[,1],
                 "time:treat_LV1" = asca0_svd$`time:treatment`$rotation[,1], 
                 "time:treat_LV2" = asca0_svd$`time:treatment`$rotation[,2]) %>% 
  add_column(species = paste0("s",1:10)) %>% 
  mutate(species = factor(species, levels = paste0("s",1:10))) %>% 
  pivot_longer(contains("LV")) %>% 
  mutate(name = factor(name)) %>% 
  mutate(name = fct_relevel(name, "treatment_LV1", after = 2)) %>% 
  ggplot() +
  geom_col(aes(x = species, y = value), fill = "steelblue") + 
  geom_hline(yintercept = 0, lty = 2) + 
  facet_wrap(~name, ncol = 1) + 
  theme_bw() + 
  theme(aspect.ratio = 0.4)

```


#### Combined Figure

```{r fig.height=9, fig.width=8, message=FALSE, warning=FALSE, dpi=250}

## Define the layout
layout <- "
AAAA
BBCC
"

## prepare the figure
screeplot + 
  LV_plot + 
  loading_plot +   
  plot_annotation(tag_levels = 'a') + 
  plot_layout(heights = unit(c(3,13),"cm"), widths =  unit(c(4,3),"cm"), design = layout)


```

## Combining terms

The function `ASCA_combine_terms` allows the combination of different factor matrices. The function accepts as input the output of `ASCA_decomposition` and a character vector specifying which terms should be combined.

```{r}
## combine two terms
asca0_comb <- ASCA_combine_terms(asca0, c("time","time:treatment"))

## perform the svd of the combined decomposition 
asca0_comb_svd <- ASCA_svd(ASCA_trim_vars(asca0_comb$decomposition, keep = 1:10))
```


### Visualize screeplot LVs and Loadings of the combined term

#### Screeplot
```{r}
## here we make a scree plot of the three decomposition terms
screeplot_comb <- asca0_comb_svd$`time+time:treatment`$sdev %>% 
  as_tibble(., rownames = "PC") %>% 
  mutate(PC = factor(PC, levels = unique(PC))) %>% 
  filter(PC %in% as.character(1:6)) %>% 
  ggplot() + 
  geom_col(aes(x = PC, y = value^2), fill = "steelblue") + 
  ylab("Variances") + 
  xlab("LV") + 
  theme_bw() + 
  theme(aspect.ratio = 0.3)

```


#### LVs

```{r}
LV_plot_comb <- cbind.data.frame("time+time:treat_LV1" = asca0_comb_svd$`time+time:treatment`$x[,1], 
                                 "time+time:treat_LV2" = asca0_comb_svd$`time+time:treatment`$x[,2],
                                 "time+time:treat_LV3" = asca0_comb_svd$`time+time:treatment`$x[,3]) %>% 
  mutate(across(everything(),~round(.x,2))) %>%     
  add_column(synth_count_data$design) %>% 
  pivot_longer(contains("LV")) %>% 
  unique() %>% 
  mutate(name = factor(name)) %>% 
  ggplot() +
  geom_point(aes(x = time, y = value, col = treatment), alpha = 0.5, size = 3) + 
  geom_line(aes(x = time, y = value, col = treatment, group = treatment, lty = treatment)) + 
  facet_wrap(~name, ncol = 1, scales = "free_y") + 
  scale_color_brewer(palette = "Set1") +
  theme_bw() + 
  theme(aspect.ratio = 0.4, legend.position = "bottom")


```

#### Loadings

```{r}
loading_plot_comb <- cbind.data.frame("time+time:treat_LV1" = asca0_comb_svd$`time+time:treatment`$rotation[,1], 
                                 "time+time:treat_LV2" = asca0_comb_svd$`time+time:treatment`$rotation[,2],
                                 "time+time:treat_LV3" = asca0_comb_svd$`time+time:treatment`$rotation[,3]) %>% 
  add_column(species = paste0("s",1:10)) %>% 
  mutate(species = factor(species, levels = paste0("s",1:10))) %>% 
  pivot_longer(contains("LV")) %>% 
  mutate(name = factor(name)) %>% 
  ggplot() +
  geom_col(aes(x = species, y = value), fill = "steelblue") + 
  geom_hline(yintercept = 0, lty = 2) + 
  facet_wrap(~name, ncol = 1) + 
  theme_bw() + 
  theme(aspect.ratio = 0.4)


```


#### combined Figure

```{r fig.height=8, fig.width=7, message=FALSE, warning=FALSE, dpi=250}

## Define the layout
layout <- "
AAAA
BBCC
"

## prepare the figure
screeplot_comb + 
  LV_plot_comb + 
  loading_plot_comb +   
  plot_annotation(tag_levels = 'a') + 
  plot_layout(heights = unit(c(4,8),"cm"), widths =  unit(c(3.5,3.5),"cm"), design = layout)


```


