---
title: "Synthetic_Dataset"
author: "Pietro Franceschi"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(reshape2)
```


## Introduction

The objective of this document is to show how the synthetic dataset included in the `gASCA` package was created.
In this simulation, 15 count variables are obtained as a function of 3 different treatments at 5 points in time. These variables could represent, for instance, count data from 10 different invertebrate species found on plants at Time (A)  0, 1, 2, 3, 4 after the plants have been treated with one of three treatments (B1, B2, B3). For each combination of time and treatment, K = 10 replicates (k = 1...K) were simulated considering the model:

$$
Y = g^{-1}(H_{0}+H_{A}+H_{B}+H_{AB}) + E_{Pois}
$$

This is further decomposed in term-specific latent variables. A set of loadings is then used to associate each latent variable to a group of the measured variables:


$$
Y = g^{-1}(H_{0}+T_{A}P^{T}_{A}+T_{B}P^{T}_{B}+T_{AB}P^{T}_{AB}) + E_{Pois}
$$

In particular.

* Two principal components are used to describe the variation in matrix $H_{A}$ (time). The variation described by the first component is only present in the first two variables, and the variation of the second component is only present in variables 3 and 4. PC1 has a continuous upward trend in time while PC2 increases in the first three time steps and then decreases again. 
* One principal component is used to describe the treatment effect in matrix $H_{B}$ (treatment) and this variation is only present in variables 5 and 6. 
* Two components are used for the interaction effect in matrix $H_{AB}$: PC1 is only active in variables 7 and 8 and shows a disagreement between treatment B3 and the other two treatments, while PC2 variation, only assigned to variables 9 and 10, represents a deviation from treatment B2 compared to the other two treatments. 
* Five additional variables not responding to the design are added to the dataset.

 
```{r}
ntime <- 5   # no of timepoints
ntreat <- 3  # no of treatments
nrepl <- 10   # no of replicates

## number of samples
nSamp = ntime * ntreat * nrepl

```


## Factor A: time

```{r}
## The design a matrix for the first factor is the following
contrA <- contr.sum(ntime)

contrA
```


For each time point we have to allocate three treatments and 10 replicates


```{r}
## since we have to allocate for each time point three treatment levels and 
## three replicates
XA <- contrA %x% rep(1,ntreat)
XA <- XA %x% rep(1,nrepl)
```

Let's now plot the matrix 

```{r}
XA %>% 
  melt() %>% 
  ggplot() + 
  geom_raster(aes(x = Var2, y = -Var1, fill = factor(value))) + 
  scale_fill_brewer(palette = "Set1", name = "Value") + 
  theme_void()
 
```

The previous design matrix is used on the two time eigentrends, which will be now organized in a matrix. Remember that since we use sum coding for the contrasts we need only four level for the 5 timepoints:

```{r}
A_LV1 <- c(-2,-1,0,1)
A_LV2 <- c(-1,0.5,1,0.5)

WA <- cbind(A_LV1,A_LV2)

```


The final score matrices are obtained combining WA and XA

```{r}
TA <- XA %*% WA
```


Let's visualize the profile of the two eigentrends


```{r}
TA %>% 
  data.frame() %>% 
  add_column(time = rep(c("t0","t1","t2","t3","t4"), each = 30)) %>% 
  add_column(treatment = rep(rep(c("B1","B2","B3"), each = 10), times = 5)) %>%  
  pivot_longer(contains("LV")) %>% 
  ggplot() + 
  geom_point(aes(x = time, y = value, col = treatment), alpha = 0.5, size = 1) + 
  geom_line(aes(x = time, y = value, col = treatment, lty = treatment, group = treatment)) + 
  facet_grid(treatment~name) + 
  theme_bw() +
  theme(aspect.ratio = 0.5)
  
```

## Factor B: treatment

As before we start from the design matrix with sum contrasts and 3 levels (B1,B2,B3)

```{r}
contrB <- contr.sum(ntreat)
## we then have n time points and the replicates
XB <- rep(1,ntime) %x% contrB
XB = XB %x% rep(1,nrepl)

```


```{r}
XB %>% 
  melt() %>% 
  ggplot() + 
  geom_raster(aes(x = Var2, y = -Var1, fill = factor(value))) + 
  scale_fill_brewer(palette = "Set1", name = "Value") + 
  theme_void()
 
```

Now we define the effect of the treatment which has only one LV

```{r}
## this is the effect of the treatment, only one component
WB <- c(-2, 1.5)
```

As before we "populate" the design matrix with the treatment levels


```{r}
TB <- XB %*% WB
```


And we plot the results

```{r}
TB %>% 
  data.frame() %>% 
  add_column(time = rep(c("t0","t1","t2","t3","t4"), each = 30)) %>% 
  add_column(treatment = rep(rep(c("B1","B2","B3"), each = 10), times = 5)) %>%  
  ggplot() + 
  geom_point(aes(x = time, y = ., col = treatment), alpha = 0.5, size = 3) + 
  geom_line(aes(x = time, y = ., col = treatment, group = treatment)) + 
  scale_color_brewer(palette = "Set1",  name = "Eigentrends") + 
  theme_bw()
```

## Interaction Term

The design matrix of the interaction term is obtained by a combination of XA and XB, each sample will be assigned to a combination of the levels of A and of B 

```{r}
## now we have the interaction term
XAB <-  cbind(XA[,1]*XB[,1], 
            XA[,1]*XB[,2], 
            XA[,2]*XB[,1], 
            XA[,2]*XB[,2], 
            XA[,3]*XB[,1], 
            XA[,3]*XB[,2], 
            XA[,4]*XB[,1], 
            XA[,4]*XB[,2])

```

and this is the plot

```{r}
XAB %>% 
  melt() %>% 
  ggplot() + 
  geom_raster(aes(x = Var2, y = -Var1, fill = factor(value))) + 
  scale_fill_brewer(palette = "Set1", name = "Value") + 
  theme_void()
 
```

```{r}
## these are the impact of the interaction on the two  components
AB_LV1 <-  3*c(0.4, 0.4, 0.2, 0.2, -0.2, -0.2, -0.4, -0.4)
AB_LV2 <-  c(-0.2,0.4, 0.2, -0.4, -0.2, 0.4, 0.2, -0.4)
WAB <- cbind(AB_LV1, AB_LV2)
```


```{r}
TAB <- XAB%*%WAB
```



```{r}
TAB %>% 
  data.frame() %>% 
  add_column(time = rep(c("t0","t1","t2","t3","t4"), each = 30)) %>% 
  add_column(treatment = rep(rep(c("B1","B2","B3"), each = 10), times = 5)) %>%  
  pivot_longer(contains("LV")) %>% 
  ggplot() + 
  #geom_point(aes(x = time, y = value, col = treatment), alpha = 0.5, size = 3) +
  geom_point(aes(x = time, y = value, col = treatment), alpha = 0.5, size = 3, 
             position=position_jitterdodge(dodge.width = 0.1, jitter.width = 0.1)) +
  
  geom_line(aes(x = time, y = value, col = treatment, lty = treatment, group = treatment),
            position=position_jitterdodge(dodge.width = 0.1, jitter.width = 0.1) ) + 
  facet_wrap(~name) + 
  scale_fill_brewer(palette = "Set1",  name = "Eigentrends") + 
  xlab("Sample Index") + 
  theme_bw() + 
  theme(aspect.ratio = 0.5)
  
```

According to the initial plan, LV1 reflects a temporal change in the difference between treatment B3 and the other treatments; LV2, instead, reflects a difference between B2 and the other two treatments which changes with time.


## Project on the species

Now we will use a set of loadings to associate the LV of each term to the 10 measured species. The loadings will be also normalized to 1. As discussed before, we also add 5 species which are not responding to the design.


```{r}
## here we use "dummy loadings" to move from the LVs to the 15 "measured" variables

## LV1 of the time factor goes on the first two species
PA1 <- c(1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
PA1 <- PA1/sqrt(sum(PA1%*%PA1)) ## normalize  

## LV2 of the time factor goes on s3 and s4
PA2 <- c(0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
PA2 <- PA2/sqrt(sum(PA2^2))

## here the loadings matrix for time
PA <- cbind(PA1, PA2)

## LV1 for the treatment factor goes to 
PB <- c(0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 0)
PB <- PB/sqrt(sum(PB^2))

## LV1 for the interaction goes to variables s7, s8
PAB1 <- c(0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0, 0, 0)
PAB1 <- PAB1/sqrt(sum(PAB1^2))
## LV2 for the interaction goes to variables s9, s10
PAB2 <- c(0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 0, 0, 0)
PAB2 <- PAB2/sqrt(sum(PAB2^2))
PAB <- cbind(PAB1, PAB2)

```


Let's plot the loadings


```{r fig.height=7, fig.width=5}
cbind.data.frame(PA,PB,PAB) %>% 
  add_column(species = paste0("s",1:15)) %>% 
  mutate(species = factor(species, levels = unique(species))) %>% 
  pivot_longer(starts_with("p")) %>% 
  mutate(name = factor(name, levels = c("PA1","PA2","PB","PAB1","PAB2"))) %>% 
  ggplot() + 
  geom_col(aes(x = species, y = value, fill = name)) +
  facet_wrap(~name, ncol = 1) + 
  theme_bw() +
  theme(aspect.ratio = 0.3)
```



Now we create the matrices containing the expected values (remember that we are working on the linear predictor space)

```{r}
A <- TA%*%t(PA)
B <- TB%*%t(PB)
AB <- TAB%*%t(PAB)
```


To reduce the number of zeroes we create a matrix with the grand mean for the different species.


```{r}
## Finally we need the grand mean
mu <- rep(1, ntreat*ntime*nrepl) %*% t(c(2, 3, 4, 5, 6, 6, 5, 4, 3, 2, 6, 5, 8, 5, 3))

```


Now we combine everything, back transform values in the "count" space and add poisson noise.


```{r}
## we add everything up and make the inverse link
logy <- mu + A + B + AB 
y <- exp(logy)

ycount <- apply(y,2, function(c) rpois(n = length(c), lambda = c))


```


To finish with the synthetic dataset we need to prepare the dataframe with the design.


```{r}
d <- data.frame(time = rep(c("t0","t1","t2","t3","t4"), each = 30),
                treatment = rep(rep(c("B1","B2","B3"), each = 10), times = 5))
colnames(ycount) <- paste0("s",1:15)


synth_count_data <- list(counts = ycount, design = d)
```


Save the data in the package folder.


```{r}
save(synth_count_data, file = "../../data/synth_count_data.RData", version = 2)
```


## Figures

We finally use `patchwork` to produce the figures for the paper.


```{r}
figure_1a <- cbind.data.frame(PA,PB,PAB) %>% 
  add_column(species = paste0("s",1:15)) %>% 
  mutate(species = factor(species, levels = unique(species))) %>% 
  pivot_longer(starts_with("p")) %>% 
  mutate(name = factor(name, levels = c("PA1","PA2","PB","PAB1","PAB2"),
                       labels = c("time_LV1", "time_LV2", "treatment_LV1", 
                                  "time:treat_LV1","time:treat_LV2" ))) %>% 
  ggplot() + 
  geom_col(aes(x = species, y = value), fill = "steelblue") +
  facet_wrap(~name, ncol = 1) + 
  ggtitle("Loadings") + 
  theme_bw() +
  theme(aspect.ratio = 0.3, 
        legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```



```{r}
figure_1b <- cbind.data.frame(TA,B_LV1 = TB,TAB) %>% 
  add_column(time = rep(c("t0","t1","t2","t3","t4"), each = 30)) %>% 
  add_column(treatment = rep(rep(c("B1","B2","B3"), each = 10), times = 5)) %>% 
  pivot_longer(contains("LV")) %>% 
  mutate(name = factor(name, levels = c("A_LV1","A_LV2","B_LV1","AB_LV1","AB_LV2"),
                       labels = c("time_LV1", "time_LV2", "treatment_LV1", 
                                  "time:treat_LV1","time:treat_LV2" ))) %>%
  ggplot() + 
  geom_point(aes(x = time, y = value, col = treatment), 
             position=position_jitterdodge(dodge.width = 0.1, jitter.width = 0.1), alpha = 0.5, size = 2) + 
  geom_line(aes(x = time, y = value, col = treatment, group = treatment, lty = treatment),
            position=position_jitterdodge(dodge.width = 0.1, jitter.width = 0.1)) + 
  facet_wrap(~name, ncol = 1) + 
  scale_color_brewer(palette = "Set1") + 
  ggtitle("LVs") + 
  theme_bw() + 
  theme(aspect.ratio = 0.3,
        legend.position = "left")
```


Join the figure with `patchwork`


```{r fig.height=7, fig.width=10}
figure_1b | figure_1a

```
















